Forest Park Trail Runs
======================
Map trail running routes through [Forest Park](http://www.portlandoregon.gov/parks/finder/index.cfm?&propertyid=127&action=ViewPark).

Last updated `r paste(Sys.time())` using `r R.version$version.string`.


Find the geographic mean of [Wildwood Trail](http://en.wikipedia.org/wiki/Forest_Park_(Portland,_Oregon)#Wildwood_Trail). Use [Yassine Diboun's](http://www.yassinediboun.com/) Wildwood Trail end-to-end-to-end fastest known time [GPS data](http://yassinediboun.com/yassinediboun/blog/item/wildwood-end-to-end-to-end-speed-record-605-miles?category_id=379)
```{r}
require(plotKML)
gpxWW <- readGPX("yassine WW double FKT.gpx")
dfWW <- gpxWW$tracks[[1]][[1]]
coordWW <- dfWW[, c("lon", "lat")]
require(geosphere)
geomeanWW <- geomean(coordWW)
geomeanWW
require(ggmap)
mapWW <- get_map(location=geomeanWW, maptype="terrain", source="stamen", zoom=12)
ggmap(mapWW)
```

Use my running history as a starting point. *Will eventually want to get other runners' data.*

Get files names of GPX data files.
```{r}
path <- file.path("..", "RunningRoutes", "activities")
files <- dir(path=path, pattern = "\\.gpx")
```
Set the date range in days. Use only the files within range of the latest file. The dates of the files are determined from the file names.
```{r}
rangeDate <- 365.25 * 3
dates <- as.Date(substr(files, 1, 8), format="%Y%m%d")
isRangeDate <- max(dates) - dates < rangeDate
files <- files[isRangeDate]
dateFrom <- min(dates[isRangeDate])
dateTo   <- max(dates[isRangeDate])
message(sprintf("Using routes run from %s to %s", dateFrom, dateTo))
```
Consolidate routes in one data frame.
```{r}
index <- c()
date <- c()
latitude <- c()
longitude <- c()
require(plotKML)
for (i in 1:length(files)) {
  f <- file.path(path, files[i])
  route <- readGPX(f)
  location <- route$tracks[[1]][[1]]
  index <- c(index, rep(i, dim(location)[1]))
  date <- c(date, rep(as.Date(substr(files[i], 1, 8), format="%Y%m%d"), dim(location)[1]))
	latitude <- c(latitude, location$lat)
	longitude <- c(longitude, location$lon)
}
routes <- data.frame(cbind(index, date, latitude, longitude))
```
Calculate distance from the central point. I'll need to use the `distHaversine` function from the [geosphere](http://www.inside-r.org/packages/cran/geosphere) package to calculate the distance from home to route coordinates. The Haversine algorithm isn't the most accurate, but it is good enough for this purpose.
```{r}
require(geosphere)
dist <- distHaversine(geomeanWW, routes[, c("longitude", "latitude")]) / 1000
routes <- data.frame(routes, dist)
```
Set the distance range in kilometers. Subset routes to those having at least one coordinate point within range from the central point.
```{r}
rangeDist <- 5
aggRoutes <- aggregate(dist ~ index, routes, min)
includedRoutes <- aggRoutes$index[aggRoutes$dist < rangeDist]
routesInRange <- subset(routes, index %in% includedRoutes)
message(sprintf("Plotting %.0d routes within %.0d km of (%.3f, %.3f)", length(includedRoutes), rangeDist, geomeanWW[1], geomeanWW[2]))
```
Set titles for map.
```{r}
fmtDate <- "%d %b %Y"
title <- sprintf("Forest Park trail runs\n%s to %s", format(dateFrom, fmtDate), format(dateTo, fmtDate))
```
Plot the routes.
```{r mapForestParkTrailRuns, dpi=240, fig.height=7, fig.width=7}
g <- ggmap(mapWW, base_layer=ggplot(routesInRange, aes(x=longitude, y=latitude, group=index)))
g <- g + geom_path(alpha=1/4, color="#7a0177")
g <- g + labs(title=title)
g
```
